Probabilistic Programming around Gaussian Processes

### 2019-10-21 abstract/paper.tex
drafted, references missing

\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{url}
\usepackage{hyperref}

\linespread{1.03}

@@

In this work, we advocate a third approach, which combines
advantages of the other two above. We present an architecture of
a Gaussian process library that provides an efficient
implementation of Gaussian process-related computations shared
by all algorithms, but lets program, virtually without
restrictions, both Gaussian process kernels and probabilistic
models of uncertainty, and combine Gaussian processes and
probabilistic models for inference using algorithms supported by
the framework. Our implementation of probabilistic
programming-enabled Gaussian processes, GoGP, is written in Go
and makes extensive use of modelling and inference capabilities
provided by Infergo~\cite{}, a Go library for differentiable
probabilistic programming. In what follows we first outline GoGP
design  and then illustrate on case studies, the use of GoGP in
a number of case studies inspired by applications.

\section{GoGP --- library for programmable GP inference}

@@

\paragraph{Barebones}

@@

@@

@@

\section{Discussion}

>>>>

\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{url}
\usepackage{hyperref}
\usepackage{listings}[language=Golang]
\usepackage{GoMono}

\lstdefinelanguage{Golang}%
  {morekeywords=[1]{package,import,func,type,struct,return,defer,panic,%
     recover,select,var,const,iota,},%
   morekeywords=[2]{string,uint,uint8,uint16,uint32,uint64,int,int8,int16,%
     int32,int64,bool,float32,float64,complex64,complex128,byte,rune,uintptr,%
     error,interface},%
   morekeywords=[3]{map,slice,make,new,nil,len,cap,copy,close,true,false,%
     delete,append,real,imag,complex,chan,},%
   morekeywords=[4]{for,break,continue,range,go,goto,switch,case,fallthrough,if,%
     else,default,},%
   morekeywords=[5]{Println,Printf,Error,Print,},%
   sensitive=true,%
   morecomment=[l]{//},%
   morecomment=[s]{/*}{*/},%
   morestring=[b]',%
   morestring=[b]",%
   morestring=[s]{`}{`},%
}

\lstset{ % add your own preferences
    xrightmargin=1pt,
    frame=single,
    basicstyle=\ttfamily\footnotesize,
    numbers=none,
    numbersep=4pt,
    showstringspaces=false,
    tabsize=4,
    language=Golang % this is it !
}

@@

In this work, we advocate a third approach, which combines
advantages of the other two above. We present an architecture of
a Gaussian process library that provides an efficient
implementation of Gaussian process-related computations shared
by all algorithms, but lets program, virtually without
restrictions, both Gaussian process kernels and probabilistic
models of uncertainty, and combine Gaussian processes and
probabilistic models for inference using algorithms supported by
the framework. Our implementation of probabilistic
programming-enabled Gaussian processes, GoGP
(\url{http://bitbucket.org/dtolpin/gogp}), is written in Go
and makes extensive use of modelling and inference capabilities
provided by Infergo~\cite{}, a Go library for differentiable
probabilistic programming. In what follows we first outline GoGP
design  and then illustrate the use of GoGP on a number of case
studies inspired by applications.

\section{GoGP --- Gaussian Process Library for Programmable Inference}

GoGP is built around a dual view on the Gaussian process: as a
stochastic process and as a probabilistic model with respect
to kernel hyperparameters and, optionally, observations.
\lstinline{GP}, the type encapsulating a Gaussian process,
provides methods \lstinline{Absorb} and \lstinline{Produce} to 
add observations to the process and predicting the distribution
of outputs at given locations. In addition, \lstinline{GP}
provides methods \lstinline{Observe} and \lstinline{Gradient}
that turn it into an Infergo model (with a custom gradient).
Inference can be performed on a \lstinline{GP} instance alone
or on a model that combines a \lstinline{GP} instance with
beliefs about hyperparameters and observations.  The
log-likelihood gradient is computed with respect to
hyperparameters only or with respect to both hyperparameters and
observations, supporting both basic and advanced usage.

Common computationally-intensive algorithms are implemented in 
methods of \lstinline{GP} and rely on efficient parallelism of
the Go language and a highly optimized library for mathematical 
computations Gonum~\cite{}. Problem-specific code comprises two
parts --- a kernel and optionally a probabilistic model --- that
are both automatically differentiated by Infergo.  As a minimum,
a \lstinline{GP} instance must be initialized with a kernel, an
object of type implementing an Infergo model and computing the
covariance between two locations. A few basic kernels are
provided in GoGP as a convenience, and can be combined in derived
kernels using virtually unrestricted Go code.

% For example, here is an RBF (Normal) kernel as implemented in
% GoGP (but could also be supplied by the library user):
% \begin{lstlisting}
% type normal struct{}
% 
% func (k normal) Observe(x []float64) float64 {
% 	return k.Cov(x[0], x[1], x[2])
% }
% func (normal) NTheta() int { return 1 }
% func (normal) Cov(l, xa, xb float64) float64 {
% 	d := (xa - xb) / l
% 	return math.Exp(-d * d / 2)
% }
% \end{lstlisting}

@@

GoGP includes case studies, illustrating, on simple examples,
common patterns of GoGP use. We briefly summarize here some
of the case studies.

\paragraph{Basic usage}

In the basic case, similar to that supported by many Gaussian
process libraries, a GP directly serves as the model for
inference on hyperparameters (or the hyperparameters can be just
fixed).

The library user specifies the kernel:
\begin{lstlisting}
type Basic struct{}
func (Basic) Observe(x []float64) float64 {
    return x[0] * kernel.Normal.Observe(x[1:])
}
func (Basic) NTheta() int { return 2 }
\end{lstlisting}
and initializes \lstinline{GP} with a kernel instance:
\begin{lstlisting}
gp := &gp.GP{
    NDim:  1,
    Simil: &Basic{},
}
\end{lstlisting}

MLE inference on hyperparameters and prediction can then be performed
through library functions.

@@

If priors on hyperparameters are to be specified, the library
user provides both the kernel and the model.  \lstinline{GP} is
initialized with the kernel, and then \lstinline{GP} and the
model are combined for inference in a derived model:
\begin{lstlisting}
type Model struct {
    gp           *gp.GP
    priors       *Priors
    gGrad, pGrad []float64
}
func (m *Model) Observe(x []float64) float64 {
    var gll, pll float64
    gll, m.gGrad = \
      m.gp.Observe(x), model.Gradient(m.gp)
    pll, m.pGrad = \
      m.priors.Observe(x),model.Gradient(m.priors)
    return gll + pll
}
func (m *Model) Gradient() []float64 {
	for i := range m.pGrad {
		m.gGrad[i] += m.pGrad[i]
	}
	return m.gGrad
}
\end{lstlisting}
In \lstinline{Model}, \lstinline{gp} holds a \lstinline{GP}
instance and \lstinline{priors} holds an instance of the model
expressing beliefs about hyperparameters. A \lstinline{Model}
instance is used for inference on hyperparameters, a
\lstinline{GP} instance --- for prediction.

@@

When observation inputs are uncertain, beliefs about inputs can
be specified, and the log-likelihood gradient can be computed
with respect to both hyperparameters and observation inputs. For
example, in a time series, one can assume that observation
inputs come from a renewal process and let the inputs move 
relative to each other. Then, forecasting can be performed 
relative to posterior observation inputs.

@@

In basic usage, the observation noise is assumed to be Gaussian.
This usage is supported by initializing \lstinline{GP} with a noise
kernel, along with a similarity kernel. When the noise is not
Gaussian, an analytical solution for posterior Gaussian process
inference does not always exist. However, non-Gaussian noise is
straightforwardly supported by GoGP through supplying a
model for beliefs on observation outputs:
\begin{lstlisting}
type Noise struct {
	Y []float64 // noisy outputs
}
func (m *Noise) Observe(x []float64) float64 {
	// Laplacian noise
	for i := range m.Y {
		ll += Expon.Logp(1/math.Exp(x[s]),
			math.Abs(m.Y[i]-x[i0+n+i]))
	}
	return ll
}
\end{lstlisting}
